---
title: "Linear Regression"
author: "Sneiba"
date: "12/1/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data(cars)
```

#Linear Regression

For this analysis, we will use the cars dataset that comes with R by default. cars is a standard built-in dataset, that makes it convenient to demonstrate linear regression in a simple and easy to understand fashion. You can access this dataset simply by typing in cars in your R console. You will find that it consists of 50 observations(rows) and 2 variables (columns) ??? dist and speed. Lets print out the first six observations here..
```{r cars}
summary(cars)
```

Before we begin building the regression model, it is a good practice to analyze and understand the variables. The graphical analysis and correlation study below will help with this.

## Graphical Analysis

The aim of this exercise is to build a simple regression model that we can use to predict Distance (dist) by establishing a statistically significant linear relationship with Speed (speed). But before jumping in to the syntax, lets try to understand these variables graphically. Typically, for each of the independent variables (predictors), the following plots are drawn to visualize the following behavior:

- Scatter Plot: Visualize the response between the predictor and response
```{r}
scatter.smooth(x=cars$speed, y=cars$dist, main="Dist ~ Speed") # Scatterplot
```
- Box Plot : To spot any outlier observations in the variable. Having outliers in your predictor can drastically affect the predictions as they can easily affect the direction/slope of the line of best fit.
```{r}
par(mfrow=c(1, 2))  # divide graph area in 2 columns
boxplot(cars$speed, main="Speed", sub=paste("Outlier rows: ", boxplot.stats(cars$speed)$out))  # box plot for 'speed'
boxplot(cars$dist, main="Distance", sub=paste("Outlier rows: ", boxplot.stats(cars$dist)$out))  # box plot for 'distance'

```
- Density Plot: Check if the response variable is close to normality

```{r}
library(e1071)
par(mfrow=c(1, 2))  # divide graph area in 2 columns
plot(density(cars$speed), main="Density Plot: Speed", ylab="Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$speed), 2)))  # density plot for 'speed'
polygon(density(cars$speed), col="red")
plot(density(cars$dist), main="Density Plot: Distance", ylab="Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$dist), 2)))  # density plot for 'dist'
polygon(density(cars$dist), col="red")
```

## Correlation

Correlation is a statistical measure that suggests the level of linear dependence between two variables, that occur in pair ??? just like what we have here in speed and dist. Correlation can take values between -1 to +1. If we observe for every instance where speed increases, the distance also increases along with it, then there is a high positive correlation between them and therefore the correlation between them will be closer to 1. The opposite is true for an inverse relationship, in which case, the correlation between the variables will be close to -1.
A value closer to 0 suggests a weak relationship between the variables. A low correlation (-0.2 < x < 0.2) probably suggests that much of variation of the response variable (Y) is unexplained by the predictor (X), in which case, we should probably look for a better explanatory variable.

```{r}
cor(cars$speed, cars$dist)
```


## Build a Linear Model

Now that we have seen the linear relationship pictorially in the scatter plot and by computing the correlation, lets see the syntax for building the linear model. The function used for building linear models is lm(). The lm() function takes in two main arguments, namely: 1. Formula 2. Data. The data is typically a data.frame and the formula is a object of class formula. But the most common convention is to write out the formula directly in place of the argument as written below.

```{r}
linearModel = lm(dist~speed, data=cars) #build linear regression model on full data
print(linearModel)
```

We have dist = -17.579 + 3.932*speed

## Linear Regression Diagnostics

Now the linear model is built and we have a formula that we can use to predict the dist value if a corresponding speed is known. Is this enough to actually use this model? NO! Before using a regression model, you have to ensure that it is statistically significant. How do you ensure this? Lets begin by printing the summary statistics for linearMod.

```{r}
summary(linearModel)
```

The summary statistics above tells us a number of things. One of them is the model p-Value (bottom last line) and the p-Value of individual predictor variables (extreme right column under ???Coefficients???). The p-Values are very important because, We can consider a linear model to be statistically significant only when both these p-Values are less that the pre-determined statistical significance level, which is ideally 0.05. This is visually interpreted by the significance stars at the end of the row. The more the stars beside the variable???s p-Value, the more significant the variable.

When the model co-efficients and standard error are known, the formula for calculating t Statistic and the p-value is as follows:
$$
  tStatistic = \frac{\beta - coefficient}{Std. Error}
$$

```{r}
modelSummary = summary(linearModel) # Capture model summary
modelCoeffs = modelSummary$coefficients #model Coefficients
beta.estimate = modelCoeffs["speed", "Estimate"] # get beta estimate for speed
std.error = modelCoeffs["speed", "Std. Error"] # get std.error for speed
t_value = beta.estimate / std.error # Calc the t_Statistic
p_value = 2*pt(-abs(t_value), df=nrow(cars)-ncol(cars)) # calc p Value
f_statistic = linearModel$fstatistic[1] #fstatistic
f = summary(linearModel)$fstatistic # parameters for model p-value calc
model_p = pf(f[1], f[2], f[3], lower=FALSE)
```

## R-Squared And Adj R-Squared

What R-Squared tells us is the proportion of variation in the dependent variable that has been explained by the model.
$$R^2 = 1-\frac{SSE}{SST}$$
where, SSE is the sum of squared errors given by $SSE = \sum^n_i(y_i-\hat{y_i})^2$ and $SST = \sum^n_i(y_i-\bar{y_i})^2$ is the sum of squared total. Here, $\hat{y_i}$ is the fitted value for observation $i$ and $\bar{y}$ is the mean of $Y$.
We don???t necessarily discard a model based on a low R-Squared value. Its a better practice to look at the AIC and prediction accuracy on validation sample when deciding on the efficiency of a model.

Now thats about R-Squared. What about adjusted R-Squared? As you add more X variables to your model, the R-Squared value of the new bigger model will always be greater than that of the smaller subset. This is because, since all the variables in the original model is also present, their contribution to explain the dependent variable will be present in the super-set as well, therefore, whatever new variable we add can only add (if not significantly) to the variation that was already explained. It is here, the adjusted R-Squared value comes to help. Adj R-Squared penalizes total value for the number of terms (read predictors) in your model. Therefore when comparing nested models, it is a good practice to look at adj-R-squared value over R-squared.

$$
  R^2_{adj} = 1 - \frac{MSE}{MST}
$$
where, MSE is the $\textit{mean squared error}$ given by $MSE = \frac{SSE}{n-q}$ and $MST = \frac{SST}{n-1}$ is the $\textit{mean squared total}$, where $n$ is the number of observations and q is the number of coefficients in the model.
Therefore, by moving around the numerators and denominators, the relationship between $R^2$ and $R^2_{adj}$ becomes:

$$
R^2_{adj} = 1 - \Bigg(\frac{(1-R^2)(n-1)}{n-q})\Bigg)
$$
## Standard Error and F-Statistic

Both standard error and F-Statistic are measures of goodness of fit

$$Std.Error = \sqrt{MSE} = \sqrt{\frac{SSE}{n-q}}$$
$$F - Statistic = \frac{MSR}{MSE}$$

where, n is the number of observations, q is the number of coefficients and $MSR$ is the $\textit{mean square regression}$, calculated as, 

$$
  MSR = \frac{\sum^n_i(\hat{y_i} - \bar{y_i})}{q-1} = \frac{SST - SSE}{q-1}
$$
## AIC and BIC

The $\textbf{Akaike's information criterion - AIC}$ and the $\textbf{Bayesian Information criterion - BIC}$ are measures of the goodness of fit of an estimated statistic model and can also be used for model selection.

For model comparison, the model with the lowest AIC and BIC score is preferred.

```{r}
AIC(linearModel)
BIC(linearModel)
```

## How to know if the model is best fit for your data ?

The most common metrics to look at while selecting the model are:

|   Statistic   |            Criterion         |
|---------------|------------------------------|
| R-Squared     | Higher the better ( > 0.70)  |
| Ajd R-Squared | Higher the better            |
| F-Statistic   | Higher the better            |
| Std.Error     | Close to zero the better     |
| t-Statistic   | Should be greater 1.96 for p-value to be less than 0.05  |
| AIC           | Lower the better             |
| BIC           | Lower the better             |

## Predicting Linear Models

So far we have seen how to build a linear regression model using the whole dataset. If we build it that way, there is no way to tell how the model will perform with new data. So the preferred practice is to split your dataset into a 80:20 sample (training:test), then, build the model on the 80% sample and then use the model thus built to predict the dependent variable on test data.

#### Step 1 : Create the training (development) and test (validation) data samples from original data.
```{r}
# Create Training and Test data -
set.seed(100)  # setting seed to reproduce results of random sampling
trainingRowIndex <- sample(1:nrow(cars), 0.8*nrow(cars))  # row indices for training data
trainingData <- cars[trainingRowIndex, ]  # model training data
testData  <- cars[-trainingRowIndex, ]   # test data
```

#### Step 2 : Develop the model on the training data and use it to predict the distance on test data

```{r}
# Build the model on training data -
lmMod <- lm(dist ~ speed, data=trainingData)  # build the model
distPred <- predict(lmMod, testData)  # predict distance
```

#### Step 3 : Review Diagnostic measures
```{r}
summary(lmMod)
```

From the model summary, the model p value and predictor???s p value are less than the significance level, so we know we have a statistically significant model. Also, the R-Sq and Adj R-Sq are comparative to the original model built on full data.

#### Step 4: Calculate prediction accuracy and error rates

A simple correlation between the actuals and predicted values can be used as a form of accuracy measure. A higher correlation accuracy implies that the actuals and predicted values have similar directional movement, i.e. when the actuals values increase the predicteds also increase and vice-versa.

```{r}
actula_preds = data.frame(cbind(actuals=testData$dist, predicteds=distPred)) # make actuals_predicteds dataframe.
correlation_accuracy = cor(actula_preds)
head(actula_preds)
correlation_accuracy
```
## k-Fold Cross Validation

Suppose the model predicts satisfactorily on the 20% split ( test data), is that enough to believe that your model will perform equally well all the time? It is important to rigorously test the model's performance as much as possible. One way is to ensure that the model equation you have will perform well, when it is 'built' on a different subset of training data and predicted on the remaining data.

How to do this is? Split your data into ???k??? mutually exclusive random sample portions. Keeping each portion as test data, we build the model on the remaining (k-1 portion) data and calculate the mean squared error of the predictions. This is done for each of the ???k??? random sample portions. Then finally, the average of these mean squared errors (for ???k??? portions) is computed. We can use this metric to compare different linear models.

By doing this, we need to check two things:
1. If the model???s prediction accuracy isn???t varying too much for any one particular sample
2. If the lines of best fit don???t vary too much with respect the the slope and level.

```{r}
install.packages("DAAG")
library(lattice)
library(DAAG)
cvResults <- suppressWarnings(CVlm(df=cars, form.lm=dist~speed, m=5, dots=FALSE, seed=29, legend.pos="topleft",  printit=FALSE, main="Small symbols are predicted values while bigger ones are actuals."));  # performs the CV
attr(cvResults, 'ms')  # => 251.2783 mean squared error
```